cardinalR: Generating interesting high-dimensional data structures

### 2025-09-19 paper-cardinalR.Rmd
Add cube functions
Updated polynomial section

### Polynomial

@@

#### `gen_quadratic()`

The `gen_quadratic(n, p, range)` function generates a dataset of $n$ points forming a quadratic curve in the first two dimensions (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:poly) a"} else { "Figure \\@ref(fig:poly-proj) a"})`). This $2\text{-}D$ parabolic structure is embedded within a higher-dimensional space with additive noise in the remaining dimensions.

The curve is constructed by randomly sampling inputs and applying a second-degree polynomial transformation. Let $X_1 \sim U(\text{range}[1], \text{range}[2])$ be the independent variable. A raw polynomial basis of degree 2 is used to form $X_2 = X_1 - X_1^2 + \varepsilon_2$, where $\varepsilon_2 \sim U(0, 0.5)$. This creates a smooth parabolic arc that opens downward, with jitter in the vertical direction to simulate noise. For $p > 2$, Gaussian noise $X_j \sim N(0, 0.1^2)$ is added to embed the $2\text{-}D$ structure into $p\text{-}D$, where $j = 3, \dots, p$.

@@

#### `gen_cubic()`

The `gen_cubic(n, p, range)` function generates a dataset of $n$ points forming a cubic curve in the first two dimensions (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:poly) b"} else { "Figure \\@ref(fig:poly-proj) b"})`). This function creates a more complex curvilinear structure than a simple parabola.

The shape is generated using a third-degree raw polynomial basis expansion. Let $X_1 \sim U(\text{range}[1], \text{range}[2])$ be the base input. A cubic transformation with vertical jitter is used to define $X_2 = X_1 + X_1^2 - X_1^3 + \varepsilon_2$, where $\varepsilon_2 \sim U(0, 0.5)$. For $p > 2$, Gaussian noise $X_j \sim N(0, 0.1^2)$ is added to embed the $2\text{-}D$ structure into $p\text{-}D$, where $j = 3, \dots, p$.

@@

### S-curve

>>>>

#### Polynomial

@@

The first is the quadratic curve of $n$ points in $p$ dimensions. This is generated using `gen_quadratic(n, p, range)`. The independent variable is defined as $X_1 \sim U(\text{range}[1], \text{range}[2])$, and a raw polynomial basis of degree 2 is applied to form $X_2 = X_1 - X_1^2 + \varepsilon_2$, where $\varepsilon_2 \sim U(0, 0.5)$. This produces a smooth parabolic arc opening downward, with vertical jitter introduced by the noise term. For $p > 2$, the quadratic structure is embedded into higher dimensions by appending additional noise coordinates  (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:poly) a"} else { "Figure \\@ref(fig:poly-proj) a"})`).

@@

The second is the cubic curve of $n$ points in $p$ dimensions. This is generated using `gen_cubic(n, p, range)`. The independent variable is defined as $X_1 \sim U(\text{range}[1], \text{range}[2])$, and a raw polynomial basis of degree $3$ is applied to construct $X_2 = X_1 + X_1^2 - X_1^3 + \varepsilon_2$, where $\varepsilon_2 \sim U(0, 0.5)$. This produces a more complex curvilinear structure than the quadratic case, with both upward and downward turning points. For $p > 2$, the cubic curve is embedded into higher dimensions by appending additional noise coordinates (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:poly) b"} else { "Figure \\@ref(fig:poly-proj) b"})`).

@@

#### S-curve

### 2025-09-19 paper-cardinalR.tex
Add cube functions
Updated polynomial section

\subsubsection{Polynomial}\label{polynomial}

@@

\paragraph{\texorpdfstring{\texttt{gen\_quadratic()}}{gen\_quadratic()}}\label{gen_quadratic}

The \texttt{gen\_quadratic(n,\ p,\ range)} function generates a dataset of \(n\) points forming a quadratic curve in the first two dimensions (Figure \ref{fig:poly-proj} a). This \(2\text{-}D\) parabolic structure is embedded within a higher-dimensional space with additive noise in the remaining dimensions.

The curve is constructed by randomly sampling inputs and applying a second-degree polynomial transformation. Let \(X_1 \sim U(\text{range}[1], \text{range}[2])\) be the independent variable. A raw polynomial basis of degree 2 is used to form \(X_2 = X_1 - X_1^2 + \varepsilon_2\), where \(\varepsilon_2 \sim U(0, 0.5)\). This creates a smooth parabolic arc that opens downward, with jitter in the vertical direction to simulate noise. For \(p > 2\), Gaussian noise \(X_j \sim N(0, 0.1^2)\) is added to embed the \(2\text{-}D\) structure into \(p\text{-}D\), where \(j = 3, \dots, p\).

@@

\paragraph{\texorpdfstring{\texttt{gen\_cubic()}}{gen\_cubic()}}\label{gen_cubic}

The \texttt{gen\_cubic(n,\ p,\ range)} function generates a dataset of \(n\) points forming a cubic curve in the first two dimensions (Figure \ref{fig:poly-proj} b). This function creates a more complex curvilinear structure than a simple parabola.

The shape is generated using a third-degree raw polynomial basis expansion. Let \(X_1 \sim U(\text{range}[1], \text{range}[2])\) be the base input. A cubic transformation with vertical jitter is used to define \(X_2 = X_1 + X_1^2 - X_1^3 + \varepsilon_2\), where \(\varepsilon_2 \sim U(0, 0.5)\). For \(p > 2\), Gaussian noise \(X_j \sim N(0, 0.1^2)\) is added to embed the \(2\text{-}D\) structure into \(p\text{-}D\), where \(j = 3, \dots, p\).

@@

\subsubsection{S-curve}\label{s-curve}

>>>>

\paragraph{Polynomial}\label{polynomial}

@@

The first is the quadratic curve of \(n\) points in \(p\) dimensions. This is generated using \texttt{gen\_quadratic(n,\ p,\ range)}. The independent variable is defined as \(X_1 \sim U(\text{range}[1], \text{range}[2])\), and a raw polynomial basis of degree 2 is applied to form \(X_2 = X_1 - X_1^2 + \varepsilon_2\), where \(\varepsilon_2 \sim U(0, 0.5)\). This produces a smooth parabolic arc opening downward, with vertical jitter introduced by the noise term. For \(p > 2\), the quadratic structure is embedded into higher dimensions by appending additional noise coordinates (Figure \ref{fig:poly-proj} a).

@@

The second is the cubic curve of \(n\) points in \(p\) dimensions. This is generated using \texttt{gen\_cubic(n,\ p,\ range)}. The independent variable is defined as \(X_1 \sim U(\text{range}[1], \text{range}[2])\), and a raw polynomial basis of degree \(3\) is applied to construct \(X_2 = X_1 + X_1^2 - X_1^3 + \varepsilon_2\), where \(\varepsilon_2 \sim U(0, 0.5)\). This produces a more complex curvilinear structure than the quadratic case, with both upward and downward turning points. For \(p > 2\), the cubic curve is embedded into higher dimensions by appending additional noise coordinates (Figure \ref{fig:poly-proj} b).

@@

\paragraph{S-curve}\label{s-curve}
